---
title: "Italian Earthquake predicion"
output: html_notebook
author: "Luca Vignali"
---

# Introduction
In this paper we utilize a model for the Italian earthquakes, and try to predict the number of earthquakes on the next day in a given area.


# The Model

# Apply the model

```{r}
library(ETAS)
library(ggmap)
library(data.table)

```


```{r}
files <- list.files(pattern = "Terremoti-201[5:6]")
temp <- lapply(files, fread)

EQCat3 <- rbindlist(temp)

# Extract only the useful information for the ETAS
# data, time, long, lat, mag

data <- strptime(EQCat3$Time, "%Y-%m-%dT%H:%M:%S", tz = "UTC")
day <- as.factor(as.Date(data))
time <- as.factor(format(data,"%H:%M"))


EQDat3 <- data.frame(date = day, time = time, long = EQCat3$Longitude, lat = EQCat3$Latitude, 
                     mag = EQCat3$Magnitude)
id <- duplicated(EQDat3[,c("date","time")])
EQDat3 <- EQDat3[!id,]



```


```{r}
fit <- readRDS("ITAFIT2002_5")
# ETASCATAll <- readRDS("ETASCATAll")



muxy <- data.frame(xi = fit$object$X$data$df$x, yi = fit$object$X$data$df$y, mu = fit$param["mu"]*fit$bk)

# This is an approximation of the expected number of earthquakes per day.
muxyV <- function(muxy,lon_min,lon_Max,lat_min,lat_Max) {
 muxy <-  muxy[muxy$xi > lon_min & muxy$xi < lon_Max,]
 muxy <- muxy[muxy$yi > lat_min & muxy$yi < lat_Max,]
 return(mean(muxy$mu)*(lon_Max - lon_min)*(lat_Max - lat_min))
}

# Let's do the same with lambda. Area 1xdegree + 1xdegree around Perugia is 12,13,42.5,43.5
library(cubature)

lambdaV <- function(lon_min,lon_Max,lat_min,lat_Max,t_min,t_max, param, hcat) {
  
  lambda2 <- function(x,param,hcat){
    lambda(x[1],x[2],x[3],param,hcat)
  }
  hcubature(lambda2,c(t_min,lon_min,lat_min),c(t_max,lon_Max,lat_Max), param, hcat)
      
}

# Find coordinates of a specific location
loc <- geocode(location = "Norcia")
lonmin <- loc$lon-0.2
lonmax <- loc$lon+0.2
latmin <- loc$lat-0.2
latmax <- loc$lat+0.2

# Calculate the prediction and the actual number of earthquakes
dateloop <- seq(as.Date(date), as.Date(date) + 10, by = 1)
prediction <- data.frame(pred = numeric(), actual = numeric())

for(i in 1:length(dateloop)) {


date <- as.POSIXct(dateloop[i])
ETASCAT <- catalog(EQDat3, mag.threshold = 2.2, flatmap = FALSE, 
                   study.end = date)
ndays <- as.numeric(date - ETASCAT$study.start)

# Number of earthquake with magnitude >= 2.2 due to background rate
a <- muxyV(muxy,lonmin,lonmax,latmin,latmax)

# Number of earthquake with magnitude >= 2.2 due to earthquakes already happened.
b <- lambdaV(lonmin,lonmax,latmin,latmax,ndays,ndays+1,fit$param,ETASCAT)

c <- catalog(EQDat3, mag.threshold = 2.2, flatmap = FALSE, 
                              study.start = date, study.length = 1,
                   lat.range = c(latmin,latmax), long.range = c(lonmin,lonmax))

p <- data.frame(pred = a + b$integral, actual = sum(c$longlat.coord$flag == 1))
               
prediction <- rbind(prediction,p)
}

# Now we should



```

# Predict 2016 Earthquake around Norcia

In this chapter we apply the model created with the 1981 - 2002 data. As the catalog 1981-2002 is stationary, we then apply the model to 2016 data. 
From the model we obtain the mu(x,y) that is background contribution and lambda(x,y,t) contribution of other earthquakes.

We focus on the area around Norcia, and want to predict the number of earthquakes with magnitude >2.2. As the contribution of the background is not relevant (0.02), we apply on

The method can be explained as follow:

1. Calculate lambda for the 1st of January 2016 based on the 2015 earthquakes catalog.
2. Store lambda.
3. Calculate lambda for 2nd of January, with catalog up to 1st January 2016 and so on.




